---
title: "Data Import Guide"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Data Import Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

molpathR provides dedicated parsers for each molecular pathology data source.
All parsers return `molpath_parsed` objects that can be combined into a unified
database using `mp_build_db()`.

## Available parsers

| Function | File types | Description |
|----------|-----------|-------------|
| `mp_read_vcf()` | .vcf, .vcf.gz | Variant Call Format files |
| `mp_read_fastq()` | .fastq, .fq | FASTQ sequence files |
| `mp_read_bam()` | .bam | BAM alignment files |
| `mp_read_xml_report()` | .xml | XML variant interpretation reports |
| `mp_read_pdf_report()` | .pdf | Pathology PDF reports |
| `mp_read_nexus_pathology()` | .csv, .xml | Nexus Pathology exports |
| `mp_read_nexus_clinical()` | .csv | Nexus clinical data exports |
| `mp_read_survival()` | .xlsx, .csv | Survival/outcome data |
| `mp_read_auto()` | any | Auto-detect and dispatch |

## Example: parsing VCF files

```{r vcf, eval = FALSE}
library(molpathR)

# Parse a single VCF file
parsed_vcf <- mp_read_vcf("path/to/variants.vcf")
parsed_vcf

# View the data
head(parsed_vcf$data)
```

## Building a database from multiple sources

```{r build, eval = FALSE}
# Parse several files
vcf1 <- mp_read_vcf("sample1.vcf")
vcf2 <- mp_read_vcf("sample2.vcf")
surv <- mp_read_survival("survival_data.xlsx")

# Build the database
db <- mp_build_db(vcf1, vcf2, surv)

# Validate integrity
mp_validate_db(db)

# Save for later use
mp_save_db(db, "my_database.rds")
```

## Auto-detection with mp_read_auto()

```{r auto, eval = FALSE}
# Automatically detects file type and uses the correct parser
parsed <- mp_read_auto("unknown_file.vcf")
parsed$source_type
```

## Handling parse errors

All parsers use defensive parsing with informative error messages:

```{r errors, eval = FALSE}
# Malformed files produce warnings, not crashes
result <- mp_read_vcf("possibly_broken.vcf")
```
