---
title: "Getting Started with molpathR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with molpathR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<!-- badges: start -->
[![R-CMD-check](https://github.com/r-heller/molpathR/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/r-heller/molpathR/actions/workflows/R-CMD-check.yaml)
[![pkgdown](https://github.com/r-heller/molpathR/actions/workflows/pkgdown.yaml/badge.svg)](https://r-heller.github.io/molpathR/)
[![CRAN status](https://www.r-pkg.org/badges/version/molpathR)](https://CRAN.R-project.org/package=molpathR)
[![Codecov test coverage](https://codecov.io/gh/r-heller/molpathR/branch/main/graph/badge.svg)](https://app.codecov.io/gh/r-heller/molpathR?branch=main)
[![CRAN downloads](https://cranlogs.r-pkg.org/badges/molpathR)](https://cran.r-project.org/package=molpathR)
[![CRAN downloads total](https://cranlogs.r-pkg.org/badges/grand-total/molpathR)](https://cran.r-project.org/package=molpathR)
[![License: MIT](https://img.shields.io/badge/license-MIT-yellow.svg)](https://opensource.org/licenses/MIT)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Installation

Install molpathR from GitHub:

```{r install, eval = FALSE}
# install.packages("remotes")
remotes::install_github("cttir/molpathR")
```

## Quick start

Load the package and create an example database:

```{r load}
library(molpathR)
db <- mp_example_db(n_patients = 50, seed = 42)
db
```

## Explore the data

Get a cohort summary:

```{r summary}
mp_summary(db)
```

## Query patients

```{r query}
# Find melanoma patients over 60
melanoma <- mp_query_patients(db, diagnosis == "Melanoma", age > 60)
head(melanoma)
```

## Query variants

```{r variants}
# Pathogenic TP53 variants with VAF > 10%
tp53 <- mp_query_variants(db, genes = "TP53", classification = "Pathogenic", min_vaf = 0.1)
head(tp53)
```

## Visualize

```{r landscape, fig.width = 8, fig.height = 5}
mp_plot_variant_landscape(db, top_n = 10)
```

```{r survival, fig.width = 7, fig.height = 5}
mp_plot_survival(db, group_by = "diagnosis", type = "os")
```

## Launch the Shiny app

```{r shiny, eval = FALSE}
mp_run_app(db)
```

# Use of LLM tools

Portions of this package were prepared with assistance from large language model tooling for
narrowly defined, non-authorial tasks: copyediting, prose smoothing, Markdown/LaTeX formatting,
scaffolding of boilerplate files (CI configs, build scripts), code refactoring. The tools used were [Chat AI](https://kisski.gwdg.de/leistungen/2-02-llm-service/),
the LLM service of KISSKI (GWDG), and a self-hosted **Mistral Small (24B, Apache-2.0)** run locally via
[Ollama](https://ollama.com/) and the `ollamar` R package — local inference only, with no data sent to
third parties for the self-hosted model.

All scientific claims, methodological choices, analyses, interpretations, and conclusions are the
author's own. No LLM-generated text was incorporated without review and revision, and every reference
was verified against its DOI, arXiv ID, or ISBN.
